Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-08-25 paper/paper.tex
update paper.tex
ave

\documentclass[onefignum, onetabnum]{siamart190516}

@@

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2021, Orus2014} is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \Cref{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \Cref{sec:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
%\purple{On a second read of this section, I think that the clarity could be improved. My understanding is this is the flow: We would like to use generic programming, where we write code that works regardless of input type, and is optimized on all input types without sacrificing efficiency. In general, we can write code such that a function runs correctly regardless of the input data type. This is easy to do for dynamically typed languages like Python, which only verify and enforce the rules for the input data type at run time (called type-checking). However, when the input data type is only checked at run time (as opposed to when the code is compiled), type-specific optimization cannot be used. High efficiency is only achieved in statically typed languages, where modern compiling technology allows the function to be individually optimized for each input data type (then give C++/Julia examples).}\blue{how about now?}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017}, the element types in the tensor networks are limited to standard number types such as floating-point numbers and integers.
%Owning to the development of modern compiling technology, we no longer need to restrict our imagination to standard number types.
%However, the elements in a tensor do not have to be a regular number type, or even a field (since division is not used in tensor contraction).
%One can use the same tensor contraction program for different purposes with different tensor element types.
% We propose to not restricting our imagination to standard number types, but to use a more general element type with certain algebraic property.
We propose to use more general element types with a certain algebraic property.
With different data types, we can solve different problems within the same unified framework.
This idea of using the same program for different purposes is also called generic programming in computer science:

@@

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Here, a commutative semiring is a field that needs not to have an additive inverse and a multiplicative inverse.
Giving up these nice properties of fields has significant implications for tensor computation: tensor network compression algorithms might not be applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014} and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
Here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we will show how to compute solution space properties of independent sets using the same tensor network contraction algorithm by engineering tensor element algebra.
The Venn diagram in \Cref{fig:venn-diagram} shows the different types of algebra we will introduce in the main text and their relation, and \Cref{tbl:generictypes} summarizes which solution space properties can be computed by which tensor element types.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To reduce the independent set problem on $G$ to a tensor network contraction, we first map a vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each vertex $v$, we define a parameterized rank-one tensor indexed by $s_v$ as
\begin{equation}
    W(x_v^{w_v}) = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with $v$.
Tensor elements can be indexed by subscripts, e.g.\ $W(x_v^{w_v})_0=1$ is the first element associated with $s_v=0$ and $W(x_v^{w_v})_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right), \label{eq:edgetensor}
\end{equation}
which encodes the independent set constraint since $B_{11} = 0$, meaning the two vertices cannot be both in the independent set.

@@

@@

    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\

@@

where $\oplus$ and $\odot$ are the standard polynomial addition and multiplication operations.

@@

\section{Maximum independent sets and its counting}\label{sec:counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
%The method we use to count the MISs is based on the following observations:
Let $x=\infty$, the independence polynomial in the previous section becomes
\begin{equation}
I(G, \infty) = \lim_{x\rightarrow \infty}\sum_{k=0}^{\alpha(G)}a_k x^k = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where all terms except the one with the largest order vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type that has two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
Here, we have generalized the previous polynomial to the Laurent polynomial to define the zero-element properly.
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
Algebra \Cref{eq:tropical} and \Cref{eq:countingtropical} are the same as those used in Liu et al.~\cite{Liu2021} to compute the spin glass ground state energy and its degeneracy.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T}(w_v) = \left(\begin{matrix}
        w_v \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may also be interested in counting the independent sets with the largest several sizes.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Cref{eq:polynomial} as:
\begin{equation}
    \eqname{P2}
    \begin{split}
    a \oplus b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
Similarly, one can also truncate the polynomial and keep only its smallest several orders.
It can be used, for example, to count the maximal independent sets with the smallest cardinality, where a maximal independent set is an independent set that cannot be made larger by adding a new vertex into it without violating the independence constraint.
As will be shown below, this algebra can also be extended to enumerate those large-size independent sets.

\section{Enumerating and sampling independent sets}\label{sec:enumeration}

\subsection{Set algebra for configuration enumeration}
The configuration enumeration of independent sets include, for example, the enumeration of all independent sets, the enumeration of all MISs, and the enumeration of independent sets with the largest several sizes.
Recall that in the definition of a vertex tensor in \Cref{eq:vertextensor}, variables carry labels, so that one can read out all independent sets directly from the output polynomials.
The multiplication between labeled variables is commutative while the summation of labeled variables forms a set.
Intuitively, one can use a bit string as the representation of a labeled variable and use the bit-wise or $\lor^\circ$ as the multiplication operation.
For example, in a 5-vertex graph, $x_2$ and $x_5$ can be represented as $01000$ and $00001$ respectively and their multiplication $x_2x_5$ can be represented as $01000 \lor^\circ 00001 = 01001$.
To enumerate all independent sets, we designed an algebra on sets of bit strings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\},
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

This set algebra can serve as the coefficients in \Cref{eq:polynomial} to enumerate independent sets of all different sizes, \Cref{eq:countingtropical} to enumerate all MISs, or \Cref{eq:max2poly} to enumerate all independent sets of sizes $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial are members of a commutative semiring, the polynomial itself is a commutative semiring.
For example, to enumerate only the MISs, we can define a combined element type $s_{k}\infty^k$, where the coefficients follow the algebra in \Cref{eq:set} and the exponents follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm P1+SN}(\{\boldsymbol{e}_{i}\}\infty^{w_v}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^{w_v}
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm P1+SN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The enumeration of all MIS configurations corresponds to the contraction of this tensor network.
However, direct contraction might have significant space overheads for keeping too many intermediate states irrelevant to the final maximum independent sets.
We introduce the bounding technique in \Cref{sec:bounding} to avoid this issue.
One may also be interested in the more studied maximal independent sets~\cite{Bron1973, Eppstein2010, Johnson1988} enumeration.
It will be discussed in \Cref{sec:maximal} since it requires using a different tensor network structure.

>>>>

\documentclass[review, onefignum, onetabnum]{siamart190516}

@@

\section{Tensor networks}\label{sec:tn}
\begin{definition}[Tensor Network~\cite{Cirac2021, Orus2014}]
    A tensor network is a triple of $\mathcal{N} = (\Tensors, \Inputs, \Output)$ where $\Tensors = (\Tensors_1, \ldots, \Tensors_m)$ is a vector of tensors, $\Inputs$ is a map from tensor to a vector of labels $\Inputs: \Tensors_j \mapsto \mathbf{l}_j$, $(\mathbf{l}_j)_i \in \Sigma, i=1,\ldots, \rm{rank}(\Tensors_j)$.
    $\Output$ is a vector of labels for specifying the output.
\end{definition}
A tensor network is composed of a collection of tensors, a collection of labels associated with tensor dimensions, and an indicator of which tensor is the output.
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network that can be represented as $C_{ik} = A_{ij}B_{jk}$, where $A, B$ and $C$ are matrices (two-dimensional tensors), $ik,ij$ and $ik$ are labels associated with them, and ``$=$'' is the indicator of $C$ being the output, while its contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \Cref{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \Cref{sec:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
%\purple{On a second read of this section, I think that the clarity could be improved. My understanding is this is the flow: We would like to use generic programming, where we write code that works regardless of input type, and is optimized on all input types without sacrificing efficiency. In general, we can write code such that a function runs correctly regardless of the input data type. This is easy to do for dynamically typed languages like Python, which only verify and enforce the rules for the input data type at run time (called type-checking). However, when the input data type is only checked at run time (as opposed to when the code is compiled), type-specific optimization cannot be used. High efficiency is only achieved in statically typed languages, where modern compiling technology allows the function to be individually optimized for each input data type (then give C++/Julia examples).}\blue{how about now?}
In previous works relating tensor networks and combinatorial problems~\cite{Kourtis2019, Biamonte2017}, the element types in the tensor networks are limited to standard number types such as floating-point numbers and integers.
%Owning to the development of modern compiling technology, we no longer need to restrict our imagination to standard number types.
%However, the elements in a tensor do not have to be a regular number type, or even a field (since division is not used in tensor contraction).
%One can use the same tensor contraction program for different purposes with different tensor element types.
% We propose to not restricting our imagination to standard number types, but to use a more general element type with certain algebraic property.
We propose to use more general element types with a certain algebraic property.
With different data types, we can solve different problems within the same unified framework.
This idea of using the same program for different purposes is also called generic programming in computer science:

@@

This motivates us to identify the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the tensor elements are members of a commutative semiring, the tensor network contraction will be well defined and the result will be independent of the contraction order.
Compared to a field, a commutative semiring does not need not to have an additive inverse and a multiplicative inverse.
Giving up these nice properties of fields has significant implications for tensor computation: tensor network compression algorithms might not be applicable because matrix factorization is NP-hard for commutative semirings~\cite{Shitov2014} and matrix multiplication faster than $O(n^3)$ does not exist for an algebra without an additive inverse~\cite{Kerr1970}.
Here, we only use the commutative properties of an algebra for optimizing the tensor network contraction order.
To define a commutative semiring with the addition operation $\OPLUS$ (not the logical \texttt{XOR} operation in computer science) and the multiplication operation $\odot$ on a set $S$, the following relations must hold for any arbitrary three elements $a, b, c \in S$.
\begin{align*}
(a \OPLUS b) \OPLUS c = a \OPLUS (b \OPLUS c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\OPLUS$ with identity $\mymathbb{0}$}\\
a \OPLUS \mymathbb{0} = \mymathbb{0} \OPLUS a = a &\\
a \OPLUS b = b \OPLUS a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\OPLUS c) = a\odot b \OPLUS a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\OPLUS b) \odot c = a\odot c \OPLUS b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
%We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we will show how to compute solution space properties of independent sets using the same tensor network contraction algorithm by engineering tensor element algebra.
The Venn diagram in \Cref{fig:venn-diagram} shows the different types of algebra we will introduce in the main text and their relation, and \Cref{tbl:generictypes} summarizes which solution space properties can be computed by which tensor element types.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To reduce the independent set problem on $G$ to a tensor network contraction, we first map each vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in $I$.
For each label $s_v$, we define a parameterized rank-one tensor indexed by it as
\begin{equation}
    W(x_v^{w_v})_{s_v} = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right)_{s_v},\label{eq:vertextensor}
\end{equation}
where $x_v$ is a variable associated with $v$.
Tensor elements can be indexed by subscripts, e.g.\ $W(x_v^{w_v})_0=1$ is the first element associated with $s_v=0$ and $W(x_v^{w_v})_1=x_v^{w_v}$ is the second element associated with $s_v=1$.
Similarly, for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right), \label{eq:edgetensor}
\end{equation}
which encodes the independent set constraint since $B_{11} = 0$, meaning the two vertices cannot be both in the independent set.

@@

\begin{theorem}
    The independence polynomial of a graph $G = (V, E)$ can be computed in time $O(|V|^22^{{\rm tw}(G)})$.
\end{theorem}

@@

    a \OPLUS b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\

@@

where $\OPLUS$ and $\odot$ are the standard polynomial addition and multiplication operations.

@@

\section{Maximum independent sets and its counting}\label{sec:counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
\begin{theorem}
    Let $G = (V, E)$ be a graph, its maximum independent set size $\alpha(G)$ can be computed in time $O(|V|2^{{\rm tw}(G)})$.
\end{theorem}
\begin{theorem}
    The number of $k$ independent sets of a graph $G = (V, E)$ can be computed in time $O(k|V|2^{{\rm tw}(G)})$.
\end{theorem}
%The method we use to count the MISs is based on the following observations:
Let $x=\infty$, the independence polynomial in the previous section becomes
\begin{equation}
I(G, \infty) = \lim_{x\rightarrow \infty}\sum_{k=0}^{\alpha(G)}a_k x^k = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where all terms except the one with the largest order vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type that has two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \OPLUS a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
Here, we have generalized the previous polynomial to the Laurent polynomial to define the zero-element properly.
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \OPLUS y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
Algebra \Cref{eq:tropical} and \Cref{eq:countingtropical} are the same as those used in Liu et al.~\cite{Liu2021} to compute the spin glass ground state energy and its degeneracy.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T}(w_v) = \left(\begin{matrix}
        w_v \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may also be interested in counting the independent sets with the largest several sizes.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Cref{eq:polynomial} as:
\begin{equation}
    \eqname{P2}
    \begin{split}
    a \OPLUS b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
Similarly, one can also truncate the polynomial and keep only its smallest several orders.
It can be used, for example, to count the maximal independent sets with the smallest cardinality, where a maximal independent set is an independent set that cannot be made larger by adding a new vertex into it without violating the independence constraint.
As will be shown below, this algebra can also be extended to enumerate those large-size independent sets.

\section{Enumerating and sampling independent sets}\label{sec:enumeration}
\begin{theorem}
    Let $G = (V, E)$ be a graph and $M$ be the number of independent sets with size $>\alpha(G)-k$.
    The independent sets with size $>\alpha(G)-k$ of $G$ can be computed in time $O(M|V|2^{{\rm tw}(G)})$.
\end{theorem}

\subsection{Set algebra for configuration enumeration}

The configuration enumeration of independent sets include, for example, the enumeration of all independent sets, the enumeration of all MISs, and the enumeration of independent sets with the largest several sizes.
Recall that in the definition of a vertex tensor in \Cref{eq:vertextensor}, variables carry labels, so that one can read out all independent sets directly from the output polynomials.
The multiplication between labeled variables is commutative while the summation of labeled variables forms a set.
Intuitively, one can use a bit string as the representation of a labeled variable and use the bit-wise or $\lor^\circ$ as the multiplication operation.
For example, in a 5-vertex graph, $x_2$ and $x_5$ can be represented as $01000$ and $00001$ respectively and their multiplication $x_2x_5$ can be represented as $01000 \lor^\circ 00001 = 01001$.
To enumerate all independent sets, we designed an algebra on sets of bit strings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \OPLUS t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\},
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \OPLUS \{01110, 01000\} = \{01110, 01000\} \OPLUS \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \OPLUS \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

This set algebra can serve as the coefficients in \Cref{eq:polynomial} to enumerate independent sets of all different sizes, \Cref{eq:countingtropical} to enumerate all MISs, or \Cref{eq:max2poly} to enumerate all independent sets of sizes $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial are members of a commutative semiring, the polynomial itself is a commutative semiring.
For example, to enumerate only the MISs, we can define a combined element type $s_{k}\infty^k$, where the coefficients follow the algebra in \Cref{eq:set} and the exponents follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \OPLUS s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm P1+SN}(\{\boldsymbol{e}_{i}\}\infty^{w_v}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^{w_v}
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm P1+SN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The enumeration of all MIS configurations corresponds to the contraction of this tensor network.
However, direct contraction might have significant space overheads for keeping too many intermediate states irrelevant to the final maximum independent sets.
We introduce the bounding technique in \Cref{sec:bounding} to avoid this issue.
One may also be interested in the more studied maximal independent sets~\cite{Bron1973, Eppstein2010, Johnson1988} enumeration.
It will be discussed in \Cref{sec:maximal} since it requires using a different tensor network structure.
